Epistemic Semantics in Guarded String Models

### 2020-09-22 tex/sdraft.tex
chopping and tightening pass

@@

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds
  models are defined, where possible worlds models are guarded string
  models in an epistemic extension of Kleene Algebra with Tests.  The
  account is framed as a formal language Epik (Epistemic KAT) for
  defining such models.  The language is interpreted by translation
  into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics
  interface for a fragment of English is defined by a categorial
  grammar.
\end{abstract}

@@

A running example of an event-sequence model is {\em The Concealed
  Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads or tails, and
neither agent knows which it is.  This initial situation is possible
world $w_1$.  Two additional worlds $w_2$ and $w_2$ are defined by
sequencing events after the initial state, with events interpreted as
in \ex{coin1}. The truth values for English sentences shown in
\ex{coin3} are observed.

@@

\begin{example} \label{coin2}
   $w_2 = w_1 \hspace{2pt} a_1 $ \hspace{2mm}
   $w_2 = w_1 \hspace{2pt} a_1 \hspace{2pt} b_1 $ 
\end{example}

@@

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between atoms, as defined in Figure
\ref{effectfigure}.  The atoms they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Aly peeking at heads) be as
specified in \ex{effect}. Then ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$
and the relation on atoms for the event $a_1$ are as given at
the bottom in \ex{effect}.

@@

\begin{figure}
\begin{center}
{\tt
\begin{tabular}{l}
state h t \\
constraint h!t + t!h \\
event a1 h:h \\
event a0 t:t \\
event b1 h:h \\
event b0 t:t \\
agent aly \\
\hspace{3mm} a1 -> a1 \\
\hspace{3mm} a0 -> a0 \\
\hspace{3mm} b1 -> b1 + b0\\
\hspace{3mm} b0 -> b1 + b0\\
agent bob \\
\hspace{3mm} b1 -> b1 \\
\hspace{3mm} b0 -> b0 \\
\hspace{3mm} a1 -> a1 + a0\\
\hspace{3mm} a0 -> a1 + a0\\
\end{tabular} }
\end{center}

@@

Guarded strings over a finite alphabet ${\sf{P}}$
are like ordinary strings, but with atoms over a set ${\sf{B}}$  
alternating with the symbols
from ${\sf{P}}$.  In the algebra described by Figure \ref{epikfigure}, 
${\sf{P}}$ is the set of events $\{a_1,a_0,b_1,b_0\}$, and
${\sf{B}}$ is $\{h,t\}$.
In the elevator example,
$\sf{P}$ is $\{u,d\}$, and $\sf{B}$ is $\{p,q\}$

Assuming a trivially true state formula $\rho$, the set of atoms
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\rho}$ in the elevator example is
$\{\bar q \bar p$, $\bar q p$, $q \bar p$, $q p\}$, which we write
$\{\hat0$, $\hat1$, $\hat2$, $\hat3\}$.  In the coin example, as we
already saw in \ex{effect}, $\mathcal{A}^\varphi_{\mathsf B}$ is
$\{h\bar{t},\bar{h}t\}$, for which we use the shorthand $\{H,T\}$.  A
guarded string over ${\sf{P}}$ and ${\sf{B}}$ is a strings of events
from ${\sf{P}}$, alternating with atoms over ${\sf{B}}$, and beginning
and ending with atoms.  \ex{gs1} gives the encoding as guarded strings
of the worlds in \ex{elevator1} and \ex{coin1}.  The length of a
guarded string $p$, written $|p|$ is the number of events in $p$.  An
atom such as $H$ is a guarded string of length $0$.

\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{clc}
{\textit{World}} & {\textit{Guarded string}} & \textit{Length}\\ \hline
%$v_1$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2$ & $4$ \\
%$v_2$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,u\, \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H \,a_1\, H$ & $1$ \\  
$w_3$ & $H\,a_1\, H \,b_1\, H$ & $2$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

\hspace{-2mm}\(H \; \diamond \; H \,a_1\, H = H \,a_0\, H \)

@@

Well-formed guarded strings have the interpretation of worlds in the
application to natural-language semantics.  The set of possible worlds
in the Kripke frame determined by an Epik specification is the set of
well-formed guarded strings.  At this point, we could say that any set
of worlds is a proposition, so that the set of propositions is the
power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead
define a more restrictive set of propositions corresponding to the
regular sets of strings. This is deferred to the next section.
Certai sets of well-formed guarded strings 
have the addional interpretation of event types.
An event-type is something that can ``happen'' is different worlds.
\ex{events1} gives the event types in
the example.

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|cl|c}
  &  {\mbox{Event type}} && {\mbox{Event type}} \\ \hline
  a_1 & \{H \,a_1\, H\} &   b_1 & \{H \,b_1\, H \} \\
  a_0 & \{T \,a_0\, T\} &   b_0 & \{T \,b_0\, T \} \\
\end{array}
\)
  \renewcommand{\arraystretch}{1}
\end{example}

@@

Consider a scenario like the one from Figure 1, but with an additional agent Cem. The
base world $T b_0 T c_0 T$ is one where the coin is tails, and first Bob looks at
tails, and them Cem looks at tails. The first event $b_0$ has the alternatives $b_0$
and $b_1$ for Amy, and the second event $c_0$ has the alternatives $c_0$ and $c_1$
for Amy.  This results in four combinations $b_0c_0$, $b_0c_1$, $b_1c_0$, and
$b_1c_1$. But these are filtered by post- and pre-conditions of events in the
alternative world, so that the set of alernatives for Amy in $T b_0 T c_0 T$ is $\{T
b_0 T c_0 T,H b_1 H c_1 H\}$, with two world-alternatives instead of four.

@@

\section{The logical language of Epistemic KAT}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  The intended interpretation of $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet *$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet \! * \! h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet\!*\!h$ is the set
of worlds where the coin is heads.\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

\begin{example}   \label{interpretation1}
\begin{tabular}[t]{ll}
${\bullet}\!* t$ & It's tails. \\
$\bullet \!* h$ & It's heads. \\  
$\Box_a \bullet \!* h$  & Amy knows that it's heads. \\
\end{tabular}
\vspace{1mm}

\begin{tabular}[t]{l}
  $\Box_b(\Box_a\!\bullet\!* t + \Box_a\!\bullet \!* \neg t)$\\
 Bob knows that Amy knows whether \\ \hspace{3mm} it's tails.  \\
\end{tabular}
\end{example}

@@

\begin{figure}
  \(\begin{array}{l >{\triangleq}c l}
  \sem{ 0 }^{\epikspec} && \emptyset \\  
  \sem{ 1 }^{\epikspec} && \mathcal A^\varphi_{\mathsf B}\\
  \sem{ b }^{\epikspec} && \mathcal A^{b\varphi}_{\mathsf B}\\
  \sem{ \bar{\sigma} }^{\epikspec} && \AphiB \setminus \sem{\sigma}^{\epikspec}\\
  \sem{ e }^{\epikspec}&&  \setabs{\alpha e \beta}{\alpha \zeta_e \beta} \\
  \sem{{p+q}}^{\epikspec} && \sem{{p}}^{\epikspec} \cup \sem{ q }^{\epikspec}\\
  \sem{{pq}}^{\epikspec} && \setabs{x \diamond y}{ \begin{array}{l}x \in \sem{p}^{\epikspec} \\ y \in \sem{q}^{\epikspec}\\\mbox{$x \diamond y$ is defined}\end{array}} \\
  \sem{p\!*\!}^{\epikspec} && \displaystyle\bigcup_{n \geq 0} {\sem{ p^n }^{\epikspec}} \\
  \sem{ \neg p }^{\epikspec} && \sem{\bullet\!*}^{\epikspec} \setminus \sem{ p }^{\epikspec} \\
  \end{array}
  \)

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a
set of guarded strings $\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be
the sets that are interpretations of terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of
epistemic KAT, with operations as in Figure \ref{fig:interp2}
(e.g. the binary operation $+$ is union, and the unary operation
$\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This provides a
concrete $n$-agent Kripke frame $\tuple{\sem{\bullet *}^{\epikspec},
  \hat{R}_1,...,\hat{R}_n}$.\footnote{ The domain of the Kripke frame
  differs from the domain of the guarded string model, because the
  former is the set of worlds, while the latter is the set of
  propositions.  } The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  It is used as a target
for natural-language interpretation in Section 6.

@@

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.  The finite state calculus has
a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has
the effect of doubling atoms at the juncture, and does not enforce
matching of atoms at the juncture.  Therefore KAT product can not be
identified with product in the finite state calculus. Instead, KAT
product and KAT Kleene star are defined operations, see Figure
\ref{fstfigure1}.  The binary product operation {\tt Cn} and the unary
Kleene star operation {\tt Kst} combine strings in the string algebra,
remove strings with non-matching atoms, and then delete the second of
two tests to create well-formed guarded strings.  Matching of atoms is
enforced with the set {\tt Wf0}, which is the set of strings that do
not contain non-matching Boolean vectors.  The containment operator
(expressed by the dollar sign) and complement (expressed by tilde) are
operators of the finite state calculus.  The set of non-matching
sequences of atoms {\tt UnequalStPair} is defined by a finite
disjunction.  Deletion is accomplished by a re-write rule in the
finite state calculus, which is a notation for defining regular
relations by contextually constrained substitutions.  In this case, is
{\tt Squash} us a regular relation that deletes an atom (a sequence of
0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state
  calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted
  in that the complement and set difference for non-equal length
  relations is not defined.  The epistemic alterative relations that
  are defined in Figure -- are equal-length relations.  }

@@

An event symbol as $a_1$ (Aly looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for each
agent. This determines a relation in the guarded string algebra a relation between
bare events decorated with compatible tests.  For agent Aly, this is the relation
described in \ex{alyrel1} as a set of ordered pairs.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).  In
the construction in Fst, the definition of world alternatives takes a
simple form. Where $R_a$ is the relation on decorated events for agent
$a$, the the corresponding relation on worlds in is the Kleene closure
of $R_a$.  Where $R$ and $S$ are relations, the concatenation product
of $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y2}$,
where $\pair{x_1}{y_1}$ is in relation $R$, and $\pair{x_2}{y_2}$ is
in relation $S$.  The Kleene closure of relation $R$ is \(\cup_{n \geq
  0}R^n\), where $R^n$ is the $n$-times concatenation product of $R$
with itself (the $0$-times concatenation product is $\sem{ 1
}^{\varphi,\mathcal E}$). This is an operation in the finite
state calculus.  Figure \ref{relclosure} defines the corresponding
operation in the guarded string algebra.  The epistemic alternative
relation on worlds for an agent is then defined as the concatenation
closure of the event alternative relation for the agent.

@@

The bounded interpretation into lists of strings is very similar to
the unbounded interpretation into sets of strings, except for the
bounds checking. The full details are shown in
Figure~\ref{fig:list-interp}. First note that when $n=0$, we simply
return the empty list, denoted $[]$. Terms of the form $0$, $1$, $e$,
and $\psi$ have the same denotation as before, translated into a list
(for a set $X$, $\lfloor X \rfloor$ is a list with the same elements
as $X$ arbitrarily ordered). The term $p+q$ denotes the list
concatenation (written $\texttt{+\!+}$) of the denotation of $p$ and of
$q$. The term $p;q$ denotes the fusion product (lifted to lists) of
the denotations of $p$ and in $q$ restricted to only those strings
shorter than $n$ (for a list $l$, $l|_n$ filters out elements longer
than $n$). The denotation of $p^*$ uses the fact that $p^*$ and $1 +
p;p^*$ are equivalent, and decrements the size threshold on the
recursive denotation of $p^*$ by $i$, where $i$ is the length of the
longest (nonzero) string in the denotation of $p$, making sure to
filter out guarded strings that are too long. The denotation of $\neg
p$ is the strings that occur in $\_^*$ and not in $p$ (the $\setminus$
operator on lists is analogous to the $\setminus$ operator on
sets). The denotation of $\Diamond_a~p$ is analogous to the set
denotation, depicted in Figure~\ref{fig:list-interp}-comprehension
notation\footnote{List comprehension notation is analogous to set
  builder notation, except that it is written using square
  brackets. The order preservation is indicated by the keyword
  \texttt{for}}.

To convert from a list of guarded strings $l$ to a set of guarded
strings, we simply write $\lceil l\rceil$. Note that for any $p$,
$\varphi$, $\mathcal E$, and $n$, $\left\lceil\llparenthesis p
\rrparenthesis^{\varphi,\mathcal E}_n\right\rceil = \{g \mid g \in
\sem{ p }^{\varphi,\mathcal E}, |g| \leq n\}$.

@@

The lists we use here are \emph{lazy} (as opposed to \emph{strict}),
which broadly means that computation is delayed until the value is
needed. This allows us to avoid computing large, unnecessary
iterations. In the elevator example, this means that $\hat 0;u;u;u^*$
will unroll the $u^*$ 0 times, once, and then discover that it is
impossible to unroll it a second time, because $\llparenthesis\hat
3;u\rrparenthesis^{\varphi,\mathcal E}_n = []$, for any $n$. So
$\llparenthesis\hat 3;u\rrparenthesis^{\varphi,\mathcal E}_n \diamond
\llparenthesis u^*\rrparenthesis^{\varphi,\mathcal E}_{n-1}$ will
return $[]$ without having to compute the full fixpoint of
$\llparenthesis u\rrparenthesis^{\varphi,\mathcal E}$. Similar
behavior occurs once $n = 0$.

Conversely, if we used sets (as in the math) instead of lists we would
need to constantly verify the set invariant (that every element is
unique) which means processing every element in the set. Verifying
that every element in $\llparenthesis
u^*\rrparenthesis^{\varphi,\mathcal E}$ is unique would require the full
computation of the fixpoint.

>>>>

\newcommand{\ehc}[1]{\textit{\textcolor{green}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}

@@

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds
  models are defined, where possible worlds models are guarded string
  models in an epistemic extension of Kleene Algebra with Tests.  The
  account is framed as a formal language Epik (Epistemic KAT) for
  defining such models.  The language is implemented by translation
  into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics
  interface for a fragment of English is defined by a categorial
  grammar.
\end{abstract}

@@

A running example of an event-sequence model is {\em The Concealed
  Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads or tails, and
neither agent knows which it is.  This initial situation is possible
world $w_1$.  Two additional worlds $w_2$ and $w_3$ are defined by
sequencing events after the initial state, with events interpreted as
in \ex{coin1}. The truth values for English sentences shown in
\ex{coin3} are observed.

@@

\begin{example} \label{coin2}
   $w_2 = w_1 \hspace{2pt} a_1 $ \hspace{2mm}
   $w_3 = w_1 \hspace{2pt} a_1 \hspace{2pt} b_1 $ 
\end{example}

@@

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between atoms, as defined in Figure
\ref{effectfigure}.  The atoms they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Amy peeking at heads) be as
specified in \ex{effect}. Then ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$
and the relation on atoms for the event $a_1$ are as given at
the bottom in \ex{effect}.

@@

\begin{figure}
  \footnotesize
\begin{center}
{\tt
\begin{tabular}{l}
state h t \\
constraint h!t + t!h \\
event a1 h:h \\
event a0 t:t \\
event b1 h:h \\
event b0 t:t \\
agent aly \\
\hspace{3mm} a1 -> a1 \\
\hspace{3mm} a0 -> a0 \\
\hspace{3mm} b1 -> b1 + b0\\
\hspace{3mm} b0 -> b1 + b0\\
agent bob \\
\hspace{3mm} b1 -> b1 \\
\hspace{3mm} b0 -> b0 \\
\hspace{3mm} a1 -> a1 + a0\\
\hspace{3mm} a0 -> a1 + a0\\
\end{tabular} }
\end{center}

@@

Guarded strings over a finite alphabet ${\sf{P}}$
are like ordinary strings, but with atoms over a set ${\sf{B}}$  
alternating with the symbols
from ${\sf{P}}$.  In the algebra described by Figure \ref{epikfigure}, 
${\sf{P}}$ is the set of events $\{a_1,a_0,b_1,b_0\}$, and
${\sf{B}}$ is $\{h,t\}$.

In the coin example, as we already saw in \ex{effect},
$\mathcal{A}^\varphi_{\mathsf B}$ is $\{h\bar{t},\bar{h}t\}$, for which we use
the shorthand $\{H,T\}$.  A guarded string over ${\sf{P}}$ and ${\sf{B}}$ is a
strings of events from ${\sf{P}}$, alternating with atoms over ${\sf{B}}$, and
beginning and ending with atoms.  \ex{gs1} gives the encoding as guarded strings
of the worlds in \ex{elevator1} and \ex{coin1}.  The length of a guarded string
$p$, written $|p|$ is the number of events in $p$.  An atom such as $H$ is a
guarded string of length $0$. Correspondingly $|w_1| = 0$, $|w_2| = 1$ and
$|w_3| = 2$.

@@

\hspace{-2mm}\(H\,b_1\,H \; \diamond \; H \,a_1\, H = H\, b_1\, H \,a_1\, H \)

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics.  The set of possible worlds in the Kripke frame
determined by an Epik specification is the set of well-formed guarded strings.
At this point, we could say that any set of worlds is a proposition, so that the
set of propositions is the power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead define a more
restrictive set of propositions corresponding to the regular sets of
strings. This is deferred to the next section.  Certain sets of well-formed
guarded strings have the addional interpretation of event types.  An event-type
is something that can ``happen'' is different worlds. For example, $a_1$ has the
event type $\{H\,a_1\,H\}$, and $a_0$ has the even type $\{T\,a_0\,T\}$. The
event types for $b_0$ and $b_1$ are analogous.

@@

Consider a scenario like the one from Figure 1, but with an additional agent Cal. The
base world $T b_0 T c_0 T$ is one where the coin is tails, and first Bob looks at
tails, and them Cal looks at tails. The first event $b_0$ has the alternatives $b_0$
and $b_1$ for Amy, and the second event $c_0$ has the alternatives $c_0$ and $c_1$
for Amy.  This results in four combinations $b_0c_0$, $b_0c_1$, $b_1c_0$, and
$b_1c_1$. But these are filtered by post- and pre-conditions of events in the
alternative world, so that the set of alernatives for Amy in $T b_0 T c_0 T$ is $\{T
b_0 T c_0 T,H b_1 H c_1 H\}$, with two world-alternatives instead of four.

@@

\section{The logical language of Epistemic KAT}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin is heads.\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

\begin{example}   \label{interpretation1}
\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
$\Box_a \bullet^* h$  & Amy knows that it's heads. \\
\end{tabular}
\vspace{1mm}

\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\bullet^*\neg t)$\\
 Bob knows that Amy knows whether \\ \hspace{3mm} it's tails.  \\
\end{tabular}
\end{example}

@@

\begin{figure}
  \(\begin{array}{l >{\triangleq}c l}
  \sem{ 0 }^{\epikspec} && \emptyset \\  
  \sem{ 1 }^{\epikspec} && \mathcal A^\varphi_{\mathsf B}\\
  \sem{ b }^{\epikspec} && \mathcal A^{b\varphi}_{\mathsf B}\\
  \sem{ \bar{\sigma} }^{\epikspec} && \AphiB \setminus \sem{\sigma}^{\epikspec}\\
  \sem{ e }^{\epikspec}&&  \setabs{\alpha e \beta}{\alpha \zeta_e \beta} \\
  \sem{{p+q}}^{\epikspec} && \sem{{p}}^{\epikspec} \cup \sem{ q }^{\epikspec}\\
  \sem{{pq}}^{\epikspec} && \setabs{x \diamond y}{ \begin{array}{l}x \in \sem{p}^{\epikspec} \\ y \in \sem{q}^{\epikspec}\\\mbox{$x \diamond y$ is defined}\end{array}} \\
  \sem{p^*}^{\epikspec} && \displaystyle\bigcup_{n \geq 0} {\sem{ p^n }^{\epikspec}} \\
  \sem{ \neg p }^{\epikspec} && \sem{\bullet^*}^{\epikspec} \setminus \sem{ p }^{\epikspec} \\
  \end{array}
  \)

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a
set of guarded strings $\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be
the sets that are interpretations of terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of
epistemic KAT, with operations as in Figure \ref{fig:interp2}
(e.g. the binary operation $+$ is union, and the unary operation
$\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This provides a
concrete $n$-agent Kripke frame $\tuple{\sem{\bullet^*}^{\epikspec},
  \hat{R}_1,...,\hat{R}_n}$.\footnote{ The domain of the Kripke frame
  differs from the domain of the guarded string model, because the
  former is the set of worlds, while the latter is the set of
  propositions.  } The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  It is used as a target
for natural-language interpretation in Section 6.

@@

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.  The finite state calculus has
a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has
the effect of doubling atoms at the juncture, and does not enforce
matching of atoms at the juncture.  Therefore KAT product can not be
identified with product in the finite state calculus. Instead, KAT
product and KAT Kleene star are defined operations, see Figure
\ref{fstfigure1}.  The binary product operation {\tt Cn} and the unary
Kleene star operation {\tt Kst} combine strings in the string algebra,
remove strings with non-matching atoms, and then delete the second of
two tests to create well-formed guarded strings.  Matching of atoms is
enforced with the set {\tt Wf0}, which is the set of strings that do
not contain non-matching Boolean vectors.  The containment operator
(expressed by the dollar sign) and complement (expressed by tilde) are
operators of the finite state calculus.  The set of non-matching
sequences of atoms {\tt UnequalStPair} is defined by a finite
disjunction.  Deletion is accomplished by a re-write rule in the
finite state calculus, which is a notation for defining regular
relations by contextually constrained substitutions.  In this case, is
{\tt Squash} is a regular relation that deletes an atom (a sequence of
0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state
  calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted
  in that the complement and set difference for non-equal length
  relations is not defined.  The epistemic alterative relations that
  are defined in Figure -- are equal-length relations.  }

@@

An event symbol such as $a_1$ (Amy looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for each
agent. This determines a relation in the guarded string algebra a relation between
bare events decorated with compatible tests.  For agent Amy, this is the relation
described in \ex{alyrel1} as a set of ordered pairs.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).  In
the construction in Fst, the definition of world alternatives takes a
simple form. Where $R_a$ is the relation on decorated events for agent
$a$, the the corresponding relation on worlds in is the Kleene closure
of $R_a$.  Where $R$ and $S$ are relations, the concatenation product
of $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$,
where $\pair{x_1}{y_1}$ is in relation $R$, and $\pair{x_2}{y_2}$ is
in relation $S$.  The Kleene closure of relation $R$ is \(\cup_{n \geq
  0}R^n\), where $R^n$ is the $n$-times concatenation product of $R$
with itself (the $0$-times concatenation product is $\sem{ 1
}^{\varphi,\mathcal E}$). This is an operation in the finite
state calculus.  Figure \ref{relclosure} defines the corresponding
operation in the guarded string algebra.  The epistemic alternative
relation on worlds for an agent is then defined as the concatenation
closure of the event alternative relation for the agent.

@@

The bounded interpretation into lists of strings is very similar to the
unbounded interpretation into sets of strings, except for the bounds
checking. The full details are shown in Figure~\ref{fig:list-interp}. First note
that when $n=0$, we simply return the empty list, denoted $[]$. Terms of the
form $0$, $1$, $e$, and $\psi$ have the same denotation as before, translated
into a list (for a set $X$, $\lfloor X \rfloor$ is a list with the same elements
as $X$ arbitrarily ordered). We compute these lists using BBDs, a standard
technique for concisely and cannoncially representing boolean functions~\cite{BBDS}.

We lift the remaining operators (except Kleene star) to their list equivalents:
union becomes list append (written $\texttt{+\!+}$); concatenation becomes the
fusion product (lifted to lists this time), negation is implemented using list
difference ($\setminus$), and the modal operator lifts the alternative relation
over lists of strings\footnote{Figure~\ref{fig:list-interp} depicts this using
  the list comprehension notation, which is analogous to set builder notation,
  except that it is written using square brackets. The order of elements is
  preserved indicated by the keyword \texttt{for}}. The only caveat to these
direct interpretations is that we restrict the operators to have size $\leq n$,
denoted as $l|_n$ for a list of guarded strings $l$.

The denotation of $p^*$ uses the fact that $p^*$ and $1 + p;p^*$ are equivalent,
and decrements the size threshold on the recursive denotation of $p^*$ by $i$,
where $i$ is the length of the longest (nonzero) string in the denotation of
$p$, making sure to filter out guarded strings that are too long.

%% To convert from a list of guarded strings $l$ to a set of guarded
%% strings, we simply write $\lceil l\rceil$. Note that for any $p$,
%% $\varphi$, $\mathcal E$, and $n$, $\left\lceil\llparenthesis p
%% \rrparenthesis^{\varphi,\mathcal E}_n\right\rceil = \{g \mid g \in
%% \sem{ p }^{\varphi,\mathcal E}, |g| \leq n\}$.

@@

The lists we use here are \emph{lazy} (as opposed to \emph{strict}), which
broadly means that computation is delayed until the value is needed. This allows
us to avoid computing large, unnecessary iterations.
